Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization

### 2020-11-14 latex/tex/algorithm.tex
tightening itemize
tightening back
fixed bug in algorithm pointed out by reviewer
updated Jupyter notebook for new plots, tighted back to 10 pages

Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
In each case, the columns of the transposed variable matrix can be computed independently.
Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
\begin{align*}
\begin{multlined}
\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
\end{multlined}
\end{align*}
where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

@@

\begin{figure}
\input{fig/split}
\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.  A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.  The 1D distribution is depicted for 3 processors to show that the splitting requires no interprocessor redistribution as the children are evenly distributed identically to the parent.}
\label{fig:split}
\end{figure}

@@

\begin{algorithm}
\caption{Parallel Rank-2 NMF}
\label{alg:parrank2nmf}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
		\State Initialize local $\M[\hat]{W}$ randomly
		\While{not converged}
			\State \Comment{Compute $\M{H}$}
			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G})$
			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
			\State \Comment{Compute $\M{W}$}
			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
		\EndWhile
	\EndFunction
	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
\end{algorithmic}
\end{algorithm}

>>>>

Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
In each case, the columns of the transposed variable matrix can be computed independently.
Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
\begin{align*}
\begin{multlined}
\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
\end{multlined}
\end{align*}
where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
We note that there are four possibilities of solutions, as each of the two variables may be positive or zero.

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the unconstrained problem does not admit a positive solution, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

@@

\begin{figure}
\input{fig/split}
\caption{Parallel splitting using Rank-2 NMF and 1D processor distribution.  A Rank-2 NMF computes factor matrices $\M{W}$ and $\M{H}$ to approximate $\M{A}$, the values of $\M{H}$ are used to determine child membership of each column (either red or blue), and the corresponding column of the $\M{W}$ matrix represents the part's feature weighting.  The 1D distribution is depicted for 3 processors to show that splitting requires no interprocessor redistribution as children are evenly distributed identically to the parent.}
\label{fig:split}
\end{figure}

@@

\begin{algorithm}
\caption{Parallel Rank-2 NMF}
\label{alg:parrank2nmf}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$ and row-distributed across processors so that $\M[\hat]{A}$ is local $(m/p)\times n$ submatrix}
	\Function{$[\M{W},\M{H}] =$ Parallel-Rank2-NMF}{$\M{A}$}
		\State Initialize local $\M[\hat]{W}$ randomly
		\While{not converged}
			\State \Comment{Compute $\M{H}$}
			\State $\M[\hat]{G}_W = \M[\hat]{W}^\Tra \M[\hat]{W}$
			\State $\M{G}_W = \textsc{All-Reduce}(\M[\hat]{G}_W)$
			\State $\M[\hat]{B} = \M[\hat]{A}^\Tra \M[\hat]{W}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{C} = \textsc{Reduce-Scatter}(\M[\hat]{B})$
			\State $\M[\hat]{H} = \textsc{Rank2-NLS-Solve}(\M[\hat]{C},\M{G}_W)$
			\State \Comment{Compute $\M{W}$}
			\State $\M[\hat]{G}_H = \M[\hat]{H}^\Tra \M[\hat]{H}$
			\State $\M{G}_H = \textsc{All-Reduce}(\M[\hat]{G}_H)$
			\State $\M{H} = \textsc{All-Gather}(\M[\hat]{H})$
			\State $\M[\hat]{D} = \M[\hat]{A} \M{H}$ %\hfill \Comment{local matrix multiplication}
			\State $\M[\hat]{W} = \textsc{Rank2-NLS-Solve}(\M[\hat]{D},\M{G}_H)$
		\EndWhile
	\EndFunction
	\Ensure{$\M{A} \approx \M{W}\M{H}^\Tra$ with $\M{W}$, $\M{H}$ row-distributed}
\end{algorithmic}
\end{algorithm}

### 2020-11-14 latex/tex/experiments.tex
updated Jupyter notebook for new plots, tighted back to 10 pages

\begin{figure}
\begin{center}
\includegraphics[height=1.4in, width=1.4in]{plots/dc_hierarchical_speedup.pdf}
\caption{Strong Scaling for Clustering on \hyper{}}
\label{fig:dcspeedup}
\end{center}
\end{figure}

@@

We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.

@@

\begin{figure}
\centering
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_rank2_speedup.pdf}
\caption{Synthetic Data}
\label{fig:synrank2speedup}
\end{center}
\end{subfigure}
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.4in]{plots/realworld_rank2_speedup.pdf}
\caption{\image{} Data}
\label{fig:rwrank2speedup}
\end{center}
\end{subfigure}
\caption{Strong Scaling Speedup for Rank-2 NMF}
\label{fig:rank2speedup}
\end{figure}

@@

\begin{figure}
\centering
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.4in]{plots/synthetic_hierarchical_speedup.pdf}
\caption{Synthetic Data}
\label{fig:synhierspeedup}
\end{center}
\end{subfigure}
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.4in]{plots/realworld_hierarchical_speedup.pdf}
\caption{\image{} Data}
\label{fig:rwhierspeedup}
\end{center}
\end{subfigure}
\caption{Strong Scaling Speedup for Clustering}
\label{fig:rank2speedup}
\end{figure}

@@

To confirm the slow growth in running time of the hierarchical algorithm in terms of the number of clusters $k$, we perform rank scaling experiments for \hyper{} and synthetic data. 
Assuming a balanced tree and relatively small $k$, \Cref{eq:treecost} shows that the dominant computational cost is proportional to $\log k$, while flat NMF algorithms have a dominant cost that is linear in $k$ \cite{KBP17}.
\Cref{fig:rankscaling} shows the raw time for various values of $k$, confirming that running time for HierNMF grows more slowly in $k$ than a flat NMF algorithm from PLANC \cite{EH+19-TR} with the same number of columns and processor grid.

\begin{figure}
\caption{\GB{To be added...}}
\label{fig:rankscaling}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/dc_hierarchical_speedup.pdf}
\caption{Strong Scaling for Clustering on \hyper{}}
\label{fig:dcspeedup}
\end{center}
\end{figure}

@@

We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and \image{} datasets.
The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and latency increases slightly with $p$.

@@

\begin{figure}
\centering
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_rank2_speedup.pdf}
\caption{Synthetic Data}
\label{fig:synrank2speedup}
\end{center}
\end{subfigure}
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/realworld_rank2_speedup.pdf}
\caption{\image{} Data}
\label{fig:rwrank2speedup}
\end{center}
\end{subfigure}
\caption{Strong Scaling Speedup for Rank-2 NMF}
\label{fig:rank2speedup}
\end{figure}

@@

\begin{figure}
\centering
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_hierarchical_speedup.pdf}
\caption{Synthetic Data}
\label{fig:synhierspeedup}
\end{center}
\end{subfigure}
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/realworld_hierarchical_speedup.pdf}
\caption{\image{} Data}
\label{fig:rwhierspeedup}
\end{center}
\end{subfigure}
\caption{Strong Scaling Speedup for Clustering}
\label{fig:rank2speedup}
\end{figure}

@@

To confirm the slow growth in running time of the hierarchical algorithm in terms of the number of clusters $k$, we perform rank scaling experiments for \hyper{} and synthetic data. 
Assuming a balanced tree and relatively small $k$, \Cref{eq:treecost} shows that the dominant computational cost is proportional to $\log k$, while a flat NMF algorithm has a dominant cost that is linear in $k$ \cite{KBP17}.
\Cref{fig:rankscaling} shows the raw time for various values of $k$, confirming that running time for HierNMF grows more slowly in $k$ than a flat NMF algorithm (based on Block Principal Pivoting) from PLANC \cite{EH+19-TR} with the same number of columns and processor grid.
We see that for sufficiently large $k$, the hierarchical algorithm outperforms flat NMF and it scales much better with $k$.

\begin{figure}
\centering
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/dc_rankscaling.pdf}
\caption{\hyper{} Data}
\label{fig:dcrankscaling}
\end{center}
\end{subfigure}
\begin{subfigure}{1.5in}
\begin{center}
\includegraphics[height=1.4in, width=1.5in]{plots/synthetic_rankscaling.pdf}
\caption{Synthetic Data}
\label{fig:synrankscaling}
\end{center}
\end{subfigure}
\caption{Rank Scaling for Hierarchical and Flat NMF}
\label{fig:rankscaling}
\end{figure}
